---
title: "Preservation: Paths"
author: "Kevin R. Coombes and Polina Bombina"
data: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Path Preservation}
  %\VignetteKeywords{DimensionReduction,Paths,Presevation}
  %\VignetteDepends{Preservation}
  %\VignettePackage{Preservation}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r opts, echo=FALSE}
knitr::opts_chunk$set(fig.width=8, fig.height=5)
```

# Introduction
**KRC: SAY SOMETHING**

As a first step, we must load the library package.
```{r packs}
library(Preservation)
```

# Preparation
## Sample Data
We need some data in order to illustrate the preservation metrics. Here we use a subset of the 
CAVA CD4 T-cell single-cell CITE-Seq data set [REF]. This data set was previously identified as
a single coherent cluster in the much larger complete data set. It contains measurements of the
expression of 51 proteins in 3096 indiciual cells. We also sub-clustered this subset into four 
parts to help with visualization.
```{r sampleData}
data("prot5")
dim(prot5)
table(colvec)
```

Next, we applied four popular dimension reduction methods:

1. Multi-Dimensional Scaling (MDS), which is equivalent to principal components analysis when
   using Euclidean distance.
2. t-Stochastic Neighbor Embedding (t-SNE).
3. Uniform Manifold Approximation and Projection (UMAP).
4. Diffusion Maps (DM).

```{r drs}
data("drs")
sapply(drs, dim)
```

Here is an initial plot showing the dimension reductions
```{r fig01, fig.width=9, fig.height=9, fig.cap="Four views of the CAVA subset after dimension reduction."}
opar <- par(mfrow = c(2,2))
for (tag in names(drs)) {
  plot(drs[[tag]], col = colvec, pch = 16, main = tag, xlab = "X1", ylab = "X2")
}
par(opar)
```

# Preservation of Pairwise Distances

```{r}
hiD <- dist(prot5)
loD <- lapply(drs, dist)
scores <- sapply(c("MilnorDistortion", "SigmaDistortion", "Stress", "SpearmanRho"), function(metric) {
  cat(metric, "\n", file = stderr())
  FUN <- get(metric)
  sapply(names(drs), function(tag) {
    FUN(hiD, loD[[tag]])
  })
})
m1d <- sapply(names(drs), function(tag) {
  M1Distortion(prot5, drs[[tag]])
})
scores <- cbind(scores, M1Distortion = m1d)
round(scores, 3)
Preservation:::PWBest(colnames(scores))
```

## First Principal Component Path
We then computed a path in high (51-)dimensional space that closely tracked the first principal
component. So, in the original space, the path is a close approximation to a line segment that
crosses the full data set. The path is represented by 301 individual cells (indicated by their
indices as rows in the data set).
```{r idex}
data("path-idex")
length(idex)
highD <- prot5[idex,]
dim(highD)
```


Here is a plot showing the dimension reductions with the overlaid image of the PC1 path.
```{r fig02, fig.width=9, fig.height=9, fig.cap="Four views of the pth of the first principal components of the CAVA subset after dimension reduction."}
opar <- par(mfrow = c(2,2))
for (tag in names(drs)) {
  lowD <- drs[[tag]][idex,]
  plot(drs[[tag]], col = colvec, pch = 16, main = tag, xlab = "X1", ylab = "X2")
  lines(lowD, lwd = 2)
}
par(opar)
```

# Path Preservation Metrics

```{r LD}
scores <- sapply(c("LengthDistortion", "SegmentVariance", "Curvature", "SpatialSimilarity"),
                 function(metric) {
  cat(metric, "\n", file = stderr())
  FUN <- get(metric)
  sapply(names(drs), function(tag) {
    lowD <- drs[[tag]][idex,]
    FUN(highD, lowD)
  })
})
round(scores, 3)

Preservation:::PathBest(colnames(scores))
```



# Appendix
This computation was perfomed using the following software tools.
```{r si}
sessionInfo()
```